---
title: "Topsis_Analysis_Caribbean_Islands"
author: "César"
date: "2022-12-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Topsis analysis of Caribean Islands

The Technique for Order of Preference by Similarity to Ideal Solution (TOPSIS) is a multiple-criteria decision making (MCDM) method.
It is based on the fundamental premise that the best solution has the shortest distance from the positive-ideal solution, and the longest distance from the negative-ideal one. Alternatives are ranked with the use of an overall index calculated based on the distances from the ideal solutions.

In this project, we employ the topsis package to perform an analysis on a dataset of different characteristics of 47 French Polynesian Caribbean islands to know which islands have the better conditions to construct a resort.

At the first part we are going to clean the data and visually explore some metrics.

## Data Cleaning and exploration


```{r}
library(topsis)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
library(forcats)
```


```{r}
github_link <- "https://raw.githubusercontent.com/cmuro27/Data_Analysis_Projects/main/dataresort.csv"
resort <- read_csv(url(github_link))
View(resort)
```

Let us have information about the dataset

```{r}
str(resort)
```
We observe that from the eighth column start the characteristics which will be evaluated for the topsis.

Let us rename the first column names
```{r}
colnames(resort)[1] <-"Island_name"
colnames(resort)[2] <-"INSEE"
colnames(resort)[4] <- "Capital"
colnames(resort)[5] <- "Area km2"
colnames(resort)[6] <- "Total_Population"
colnames(resort)[7] <- "Density"
```

Now, for the columns Total_Population and Density_(pers/km2) , we change the type as numeric. But before, we need to remove the spaces

```{r}
resort$Total_Population <- gsub('\\s+', '', resort$Total_Population)
resort$`Density` <- gsub('\\s+', '', resort$`Density`)
```
Proceedind to convert to numeric Total_Population 

```{r}
resort$Total_Population <- as.numeric(resort$Total_Population)
resort$`Density` <- as.numeric(gsub(",","",resort$`Density`))
```

Then, our dataframe takes the form

```{r}
View(resort)
```

with the characteristics 
 
 
```{r}
str(resort)
```
Let us visualize explore which islands are the biggest, and which have more density.
A bar plot by population

```{r}
resort %>% ggplot(aes(x=Total_Population,y=Island_name))+geom_bar(stat = "identity") 
```
Let us get a table with the 5 most populated islands

```{r}
 head(resort[order(resort$Total_Population,decreasing = TRUE),],5)
```

A bar plot by density
```{r}
ggplot(resort,aes(x=Density,y=Island_name))+geom_bar(stat = "identity")
```


Let us get a table with the most densely populated islands
```{r}
 head(resort[order(resort$Density,decreasing = TRUE),],5)
```

And finally, a bar plot for area

```{r}
sort_data <- resort[order(resort$`Area km2`,decreasing = TRUE),]
View(sort_data)
```


```{r}
ggplot(sort_data,aes(x=`Area km2`,y=Island_name))+geom_bar(stat = "identity")
```
 

```{r}
 head(resort[order(resort$`Area km2`,decreasing = TRUE),],5)
```

## Topsis Analysis.

We need to define a weight vector, and a sign vector that defines the sign of the influence. 
Notice that the sum of the weights must be equal to 1.
Then we use the topsis function

```{r}
w <- c(0.11,0.11,0.05,0.11,0.11,0.07,0.07,0.09,0.07,0.11,0.1) 
i <- c("+","+","+","+","+","+","+","+","+","+","+")
sum(w)

datare <- select(resort,8:18)
nm <- data.matrix(datare)
View(resort[,1])
results <- topsis(nm,w,i)
results <- data.matrix(results)
class(results)
final <- cbind.data.frame(resort[,1],results)
```
Therefore, the rank of the best islands to consider establish a resort is

```{r}
head(final[order(final$`rank`,decreasing = FALSE),],5)
```



